---
title: "Bellabeat Case Study"
author: "Rachel Zhang"
date: "2023-08-02"
output: github_document
---


## **Introduction**
   
   > This case study aims to use public data on smart device users' daily habits in order to discover consumer trends and advise *Bellabeat*'s marketing strategy.

   Founded in 2013 by Urška Sršen and Sando Mur, *Bellabeat* is a high-tech company that manufactures health-focused smart products for women. Some of their products include:
   
   * **Bellabeat App**: The Bellabeat app provides users with health data related to their activity, sleep, stress, menstrual cycle, and mindfulness habits.
   * **Leaf**: Bellabeat’s classic wellness tracker can be worn as a bracelet, necklace, or clip. The Leaf tracker connects to the Bellabeat app to track activity, sleep, and stress.
   * **Time**: This wellness watch combines the timeless look of a classic timepiece with smart technology to track user activity, sleep, and stress. The Time watch connects to the Bellabeat app to provide you with insights into your daily wellness.
   * **Spring**: This is a water bottle that tracks daily water intake using smart technology to ensure that you are appropriately hydrated throughout the day.
   
   
      
   
   The dataset consists of personal tracker data from 30 eligible Fitbit users, which includes minute-level output for physical activity, heart rate, and sleep monitoring. Following the analysis of the data, I will use my findings to provide a few recommendations for Bellabeat's products and marketing.

#### I will center this analysis on data from three spreadsheets within the dataset: daily activity, daily sleep, and hourly step count.
```{r, message=FALSE}
# Import necessary packages

library(tidyverse) #Includes ggplot2, dplyr, readr, tibble, tidyr, stringr
library(lubridate) #For formatting dates
library(RColorBrewer) #For color palettes in graphs
```
   
```{r,message=FALSE}
# Import all csv files

activity_df <- read_csv("dailyActivity_merged.csv") #daily activity in minutes
hourly_steps_df <- read_csv("hourlySteps_merged.csv") #hourly step count
sleep_df <- read_csv("sleepDay_merged.csv") #daily sleep in minutes
```

```{r}
# Ensure that data has been imported correctly:

head(activity_df)
head(hourly_steps_df)
head(sleep_df)
```
## **Data Cleaning and Formatting**
```{r}
# Nice! Now let's start cleaning. Let's check for duplicates and na values from our dataframes
# First check for na values:

any(is.na(activity_df))
any(is.na(hourly_steps_df))
any(is.na(sleep_df))
```
```{r}
# Great! No na values for any of them. 
# Now check for duplicates:

any(duplicated(activity_df))
any(duplicated(hourly_steps_df))
any(duplicated(sleep_df))
```
```{r}
# Drop the duplicates from sleep file:

sleep_df <- sleep_df %>%
  distinct() %>%
  drop_na()

# Verify that duplicates are removed:
any(duplicated(sleep_df))
```
```{r}
# For the purpose of later analyses, I'll create some new variables within the dataframes

# First, within the hourly steps dataframe, I'll make a new variable that stores the weekday from each date, and another variable that stores the hour

hourly_steps_df$weekday <- wday(mdy_hms(hourly_steps_df$ActivityHour), label=TRUE)#stores weekday from the date log
hourly_steps_df$hour <- hour(mdy_hms(hourly_steps_df$ActivityHour))

head(hourly_steps_df)
```
```{r}
# Next, I'll also create another weekday variable in the sleep and activity dataframes, and change the current date variable to show only date and no time since the time is the same for every entry

sleep_df$SleepDay <- as.Date(mdy_hms(sleep_df$SleepDay))#convert date to just month day year
sleep_df$weekday <- wday(sleep_df$SleepDay, label = TRUE)

activity_df$ActivityDate<- as.Date(activity_df$ActivityDate, format = "%m/%d/%Y")#changes activity date to year month day format 
activity_df$weekday <- wday(activity_df$ActivityDate, label = TRUE)
```

```{r}
# I'll also make a new dataframe for one of my analyses that needs variables from two dataframes
# Perform an inner join on the sleep and activity dataframes. 
sleep_df <- rename(sleep_df, Date = SleepDay)
activity_df <- rename(activity_df, Date = ActivityDate)

merged_activity_sleep <- merge(sleep_df, activity_df, by=c("Id","Date")) 
```

## **Analysis**

#### **Weekday and Total Minutes Sleeping**
   
   First, let's examine the relationship between weekday and total minutes spent sleeping through a group boxplot
```{r}
weekday_sleep_plot <- ggplot(sleep_df, aes(x = weekday, y = TotalMinutesAsleep))
weekday_sleep_plot + geom_boxplot(aes(fill = weekday))+scale_fill_brewer(palette="Set3")+labs(title = "Weekday and Total Minutes Sleeping", x = "Weekday", y = "Total Minutes Asleep") 
```
   
   From the visualization, we can see that Sunday has the highest median minutes asleep, while Friday has the lowest average minutes asleep. Saturday and Sunday also have the largest ranges of sleep time, ranging from under 200 minutes(3 hours) to over 600 minutes(10 hours). 
   This clearly shows how work schedules can have a significant effect on sleep. People tend to maintain consistent sleep on work days (Mon-Fri) with low variation, and then either spent weekends sleeping significantly more(likely "catching up on sleep") or less for nighttime activities. 
   Let's also look at the average total minutes sleeping for each weekday:
```{r}
sleep_summary <- sleep_df %>%
  group_by(weekday) %>%
  summarise(mean_value = mean(TotalMinutesAsleep,na.rm=TRUE))

sleep_summary
```
   
   The CDC recommends that adults 18-60 years sleep at least 7 hours per night. However, it appears that from our dataset the individuals do not get an average of at least 7 hours of sleep on any days except Wednesday and Sunday.

#### **Weekday vs. Total Steps**

   Similar to the first plot, a group boxplot is helpful to see the distribution of total steps for each weekday.
```{r}
weekday_step_plot <- ggplot(activity_df, aes(x = weekday, y = TotalSteps))
weekday_step_plot + geom_boxplot(aes(fill = weekday))+scale_fill_brewer(palette="Set2")+labs(title = "Weekday and Total Step Count", x = "Weekday", y = "Total Steps")

```
   
   Sunday has the smallest median for total steps. This corresponds to what we found in the other plot: because people tend to sleep more on Sundays, it can cause lower overall step count. Saturdays also appear to have the highest variance, and we can confirm that different weekend activities may cause differing activity levels. 
   
   We can also look at the average total steps for each weekday:
```{r}
step_summary <- activity_df %>%
  group_by(weekday) %>%
  summarise(mean_value = mean(TotalSteps,na.rm=TRUE))

step_summary
```
   It is recommended that people walk at least 10,000 steps per day, however none of the weekdays have an average step count that reaches 10,000.
   
#### **Total Minutes Sleeping and Total Minutes Sedentary**
   
   Next, let's look at the relationship between two variables from our merged sleep and activity dataframe: total minutes sleeping and total minutes sedentary(seated, inactive).
   
```{r, message = FALSE}
sleeping_sedentary_plot <- ggplot(merged_activity_sleep, aes(x = TotalMinutesAsleep, y = SedentaryMinutes))
sleeping_sedentary_plot + geom_point() + geom_smooth(method = "lm")+labs(title = "Total Minutes Sleeping and Total Minutes Sedentary", x = "Total Minutes Asleep", y = "Total Minutes Sedentary")
```
```{r}
cor(merged_activity_sleep$TotalMinutesAsleep,merged_activity_sleep$SedentaryMinutes) #calculate correlation coefficient
```
There appears to be a moderately strong negative correlation between total minutes asleep and total minutes sedentary. As people spend more time asleep, they tend to spend less time sedentary throughout the day.

#### **Hour of Day and Average Number of Steps**
   
   This time, we'll use a group of bar charts to look at average number of steps per hour, for each weekday.
```{r, message = FALSE}
df_avg <- hourly_steps_df %>%
  group_by(weekday, hour) %>%
  summarise(avg_steps = mean(StepTotal, na.rm = TRUE))

# Create the bar chart
ggplot(df_avg, aes(x = hour, y = avg_steps, fill = weekday)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ weekday) +
  labs(x = "Hour of the Day", y = "Average Steps", title = "Average Steps per Hour by Weekday")
```

   
   From this distribution, because the average steps for 8pm-6am are significantly less than the other hours, let's narrow down the hours displayed in the bar charts to get a closer look into the more active hours. 
```{r, message = FALSE}
df_filtered <- hourly_steps_df %>%
  filter(hour>6 & hour<20)
df_avg <- df_filtered %>%
  group_by(weekday, hour) %>%
  summarise(avg_steps = mean(StepTotal, na.rm = TRUE))

# Create the bar chart
ggplot(df_avg, aes(x = hour, y = avg_steps, fill = weekday)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ weekday) +
  labs(x = "Hour of the Day", y = "Average Steps", title = "Average Steps per Hour by Weekday, 7AM to 7PM")
```  
   
   Overall, we can see that people have the most steps per hour on Saturdays, while for other weekdays step count tends to be a lot lower. There tends to be an increase in average step count 11AM-2PM, which could be due to people moving around for lunch time, and 5-7PM for dinner time. 
   
## **Summary**   
From our analyses, we discovered some important findings:   
 
   * For each weekday, participants had an average of 6.68-7.53 hours of sleep. For only two of those weekdays did the average surpass the recommended minimum hours of sleep. 
       + Weekends also had much more variation in sleep time than work days
   * The average number of steps for each weekday ranged from approximately 6,993 to 8,152, none of which reached the recommended daily step count of 10,000
   * There appeared to be a moderately strong correlation between amount of sleep and sedentary minutes, where the more sleep people got, they tended to have less sedentary minutes.
      + Less sedentary minutes means more active minutes during the day
   *  Throughout the day, step count peaks around mealtime hours and tend to be noticeably lower outside of those hours.
   
## **Recommendations**

Based on these results from Fitbit users, I have many recommendations for Bellabeat.

   * Include short articles within their Bellabeat app that inform users of the importance of getting enough sleep and exercise every day
      + Articles can be backed by reliable sources such as the CDC
      + Can help to motivate users to adopt healthier habits
      
   * Include a function that allows users to set their desired sleep schedules and daily fitness goals.
   
      + Within those schedules, provide recommended changes to those schedules for a healthier lifestyle
         + Example: User consistently gets 5 hours of sleep, provide recommendation to try and get more sleep and pair with science-backed articles to emphasize importance of adequate sleep
   * Add a habit tracker function
      + Allows users to see overall progress in maintaining habits, can encourage them to keep it up
   * Develop a notification system connected to the Leaf and Time devices, where the physically worn accessories can vibrate to keep users on track
         + People can be less active during working hours, device can buzz every hour to remind users to get up and walk around
         + Remind users to hydrate if they haven't drank enough water
         + Remind users of bedtime, include early reminder so they can get other tasks done
   * Marketing recommendations:
      + Provide customer testimonies on improvement of lifestyle and healthier habits
      + Encourage reviews from all customers and feedback forms
      + Try to partner with large corporations to work with those that have office jobs and need to be more active
         + Can encourage company to provide incentives if workers maintain healthy habits through products
         
## **Sources:**
https://www.kaggle.com/datasets/arashnic/fitbit   
https://www.medicalnewstoday.com/articles/how-many-steps-should-you-take-a-day   
https://www.cdc.gov/sleep/about_sleep/how_much_sleep.html   


      

